Sparse Function-space Representation \\ of Neural Networks

### 2023-05-17 paper/main.tex
update intro

%Need for adaptive learning methods + failures with current methods
Given an approximate inference technique, we demonstrate that the neural network emits \todo{is 'emits' a clear enough expression here?}`dual' parameters which are the building blocks of a GP~\cite{csato2002sparse, adam2021dual, chang2020fast} . In contrast to previous work that utilizes subsets of training data \cite{immer2021scalable},
this parameterization allows capturing the contributions from {\em all} data points into a sparse representation, essential for predictive uncertainty. Crucially, the resulting GP now predicts in the same space as the original trained neural network \todo{Could make it clearer that the GP predicts in the output space while avoiding the NN parameter space}, a feature not present in previous approaches, while avoiding the notorious cubic complexity of vanilla GPs. Through the dual parameter formulation, we establish a connection between the neural network, full GPs, and a sparse approximation similar to sparse variational GPs~\cite{titsias2009variational}. We refer to our method as Sparse Function-space Representation (\our)---a sparse GP derived from a trained neural network. Moreover, this dual parameterization can be exploited to perform dual conditioning \citep{chang2022fantasizing}, \ie, an effective approach to condition on new incoming data into our model crucially without need of retraning it(see \cref{fig:teaser}).
%As \our is in the dual parameters, we can perform dual conditioning recently shown effective in \cite{chang2022fantasizing}, that is, avoid retraining and condition new data into our model (see \cref{fig:teaser}) \todo{Effective compared to what?}.
\looseness-2

@@

The contributions of this paper are:
%
{\em (i)}~We introduce \our, a new approach for building a sparse functional representation of a neural network.
{\em (ii)}~We demonstrate that, despite its sparsity, our method effectively captures predictive uncertainty, provides means of updating the model post-training, and gives a a compact regularizer suitable for continual learning.
{\em (iii)}~We provide extensive experiments for showcasing our approach and demonstrate significance and applicability across supervised, continual, and reinforcement learning, aiming to stimulate future use of the approach.

@@

% Introduce the laplace GGN + shortcomings
\textbf{Function-space methods}
Function-space perspectives on uncertainty in neural networks often use a Laplace-GGN approximation \citep{daxberger2021laplace}, that linearize a trained neural network around MAP weights and approximates the neural network's Hessian using the generalized Gauss--Newton approximation (GGN, \cite{botev2017practical}). 
%\todo{Could modify the sentence a bit, maybe "The Laplace-GGN approximates the Hessian by using..."}
%The Hessian can be efficiently approximated using the generalized Gauss--Newton approximation . 
While efficient, it suffers from cubic scaling in parameter count, necessitating approximations like Kronecker factorisation \cite{martens2015optimizing, ritter2018kfac}. \todo{Reads a bit repetitive to start two sentences in a row by This, could maybe be combined in one.} This linear model (linear in the weights) refines predictions and provides uncertainty estimates \citep{immer2021scalable}. This model has a convenient interpretation as a GP \cite{immer2021scalable, khan2019approximate, maddox2021fast}.
%
However, the GP's cubic $\mathcal{O}(n^3)$ scaling with data points $n$ requires costly approximations, often resorting to using data subsets \cite{immer2021scalable}. In the GP community, sparse approximations have mitigated this scaling issue (\eg, \cite{titsias2009variational,hensman2013gaussian}), but combining neural network linearization with sparse methods remains \todo{Maybe unexplored instead of unclear?} unclear without resorting to subset methods or separately retraining a GP model \cite{ortega2023variational} \todo{Could add something on subset methods here, for example 'which do not retain information from the full dataset' or something like that}. Our work addresses this by utilizing the GP's dual parameters \cite{csato2002sparse}, previously applied to non-conjugate likelihood models \cite{adam2021dual}.

@@

\section{\our: Sparse function-space representation of neural networks}
%\paragraph{GP in the dual parameters}
The seminal work by \citet{csato2002sparse} (parameterization Lemma~1) gives a parameterization for the posterior process that can be found through the Bayesian update using the GP prior and the likelihood function. This gives a `dual' parameterization, $\valpha$ and $\vbeta$,
%
\begin{equation}  \label{eq:gp_pred}
  \myexpect_{p(f_i \mid\vy)}[f_i]= \vk_{\vx i}^\top \valpha \quad \text{and} \quad
  \mathrm{Var}_{p(f_i \mid \vy)}[f_i] = k_{ii} - \vk_{\vx i}^\top ( \MKxx + \diag(\vbeta)^{-1})^{-1} \vk_{\vx i},
\end{equation}
%
where the $ij$\textsuperscript{th} entry of the matrix $\MKxx \in \R^{N \times N}$ is $\kappa(\vx_i,\vx_j)$, $\vk_{\vx i}$ denotes a vector where each $j$\textsuperscript{th} element is $\kappa(\vx_i, \vx_j)$, and $k_{ii} = \kappa(\vx_i, \vx_i)$. \cref{eq:gp_pred} states that the resultant posterior process, which may not be a GP, first two moments can be parameterized via the dual parameters, which are the vectors $\valpha, \vbeta \in \R^{N}$, defined as:\looseness-1
%
\begin{equation}
  \label{eq:dual_param}
  \alpha_i \coloneqq \myexpect_{p(\vw \mid \vy)}[\nabla_{f}\log p(y_i \mid f) |_{f=f_i}]
  \quad \text{and} \quad
  \beta_i \coloneqq - \myexpect_{p(\vw \mid \vy)}[\nabla^2_{f f}\log p(y_i \mid f_i) |_{f=f_i}],
\end{equation}
%
where $f_\vw(\vx_i) = f_i$ is the function output at input $\vx_i$. The relationships specified are valid for generic likelihoods, and involve no approximations since the expectation is under the exact posterior, but given that the model can be expressed in a kernel formulation. \cref{eq:gp_pred} and \cref{eq:dual_param} highlight that the approximate inference technique, usually viewed as a posterior approximation, can be alternatively interpreted as an approximation of the expectation of gradients of loss/likelihoods.

>>>>

%Need for adaptive learning methods + failures with current methods
Given an approximate inference technique, we demonstrate that the neural network emits \todo{is 'emits' a clear enough expression here?}`dual' parameters which are the building blocks of a GP~\cite{csato2002sparse, adam2021dual, chang2020fast} . In contrast to previous work that utilizes subsets of training data \cite{immer2021scalable},
this parameterization allows capturing the contributions from {\em all} data points into a sparse representation, essential for predictive uncertainty. Crucially, the resulting GP directly predicts in the same space as the original trained neural network, with the benefit of avoiding the complexity introduced by working in weight-space and the notorious cubic complexity of vanilla GPs.
%\todo{Could make it clearer that the GP predicts in the output space while avoiding the NN parameter space}
%, a feature not present in previous approaches, while avoiding the notorious cubic complexity of vanilla GPs. 
Through the dual parameter formulation, we establish a connection between the neural network, full GPs, and a sparse approximation similar to sparse variational GPs~\cite{titsias2009variational}. We refer to our method as Sparse Function-space Representation (\our)---a sparse GP derived from a trained neural network. Moreover, this dual parameterization can be exploited to perform dual conditioning \citep{chang2022fantasizing}, \ie, an effective approach to condition on new incoming data into our model crucially without need of retraning it(see \cref{fig:teaser}).
%As \our is in the dual parameters, we can perform dual conditioning recently shown effective in \cite{chang2022fantasizing}, that is, avoid retraining and condition new data into our model (see \cref{fig:teaser}) \todo{Effective compared to what?}.
\looseness-2

@@

The contributions of this paper are that:
%
{\em (i)}~We introduce \our, a new approach for building a sparse functional representation of a neural network.
{\em (ii)}~We demonstrate that, despite its sparsity, our method effectively captures predictive uncertainty, provides means of updating the model post-training, and gives a a compact regularizer suitable for continual learning.
{\em (iii)}~We provide extensive experiments for showcasing our approach and demonstrate significance and applicability across supervised, continual, and reinforcement learning, aiming to stimulate future use of the approach.

@@

% Introduce the laplace GGN + shortcomings
\textbf{Function-space methods}
Function-space perspectives on uncertainty in neural networks often use a Laplace-GGN approximation \citep{daxberger2021laplace}, that linearize a trained neural network around MAP weights and approximates the neural network's Hessian using the generalized Gauss--Newton approximation (GGN, \cite{botev2017practical}). 
%\todo{Could modify the sentence a bit, maybe "The Laplace-GGN approximates the Hessian by using..."}
%The Hessian can be efficiently approximated using the generalized Gauss--Newton approximation . 
While efficient, it suffers from cubic scaling in parameter count, necessitating approximations like Kronecker factorisation \cite{martens2015optimizing, ritter2018kfac}. \todo{Reads a bit repetitive to start two sentences in a row by This, could maybe be combined in one.} This linear model (linear in the weights) refines predictions and provides uncertainty estimates \citep{immer2021scalable}. This model has a convenient interpretation as a GP \cite{immer2021scalable, khan2019approximate, maddox2021fast}.
%
However, the GP's cubic $\mathcal{O}(n^3)$ scaling with data points $n$ requires costly approximations, often resorting to using data subsets \cite{immer2021scalable}. In the GP community, sparse approximations have mitigated this scaling issue (\eg, \cite{titsias2009variational,hensman2013gaussian}), but combining neural network linearization with sparse methods remains \todo{Maybe unexplored instead of unclear?} unclear without resorting to subset methods \todo{Could add something on subset methods here, for example 'which do not retain information from the full dataset' or something like that}. Our work addresses this by utilizing the GP's dual parameters \cite{csato2002sparse}, previously applied to non-conjugate likelihood models \cite{adam2021dual}.

@@

\section{\our: Sparse function-space representation of neural networks}
%\paragraph{GP in the dual parameters}
The seminal work by \citet{csato2002sparse} (parameterization Lemma~1) gives a parameterization for the posterior process $f_{\textrm{post}}$ \todo{The f post doesn't match the notation in Background?} that can be found through the Bayesian update using the GP prior and the likelihood function. This gives a `dual' parameterization, $\valpha$ and $\vbeta$,
%
\begin{equation}  \label{eq:gp_pred}
  \myexpect_{p(f_i \mid\vy)}[f_i]= \vk_{\vx i}^\top \valpha \quad \text{and} \quad
  \mathrm{Var}_{p(f_i \mid \vy)}[f_i] = k_{ii} - \vk_{\vx i}^\top ( \MKxx + \diag(\vbeta)^{-1})^{-1} \vk_{\vx i},
\end{equation}
%
where the $ij$\textsuperscript{th} entry of the matrix $\MKxx \in \R^{N \times N}$ is $\kappa(\vx_i,\vx_j)$, $\vk_{\vx i}$ denotes a vector where each $j$\textsuperscript{th} element is $\kappa(\vx_i, \vx_j)$, and $k_{ii} = \kappa(\vx_i, \vx_i)$.  The dual parameters are the vectors $\valpha, \vbeta \in \R^{N}$, defined as:\looseness-1
%
\begin{equation}
  \label{eq:dual_param}
  \alpha_i \coloneqq \myexpect_{p(\vw \mid \vy)}[\nabla_{f}\log p(y_i \mid f) |_{f=f_i}]
  \quad \text{and} \quad
  \beta_i \coloneqq - \myexpect_{p(\vw \mid \vy)}[\nabla^2_{f f}\log p(y_i \mid f_i) |_{f=f_i}],
\end{equation}
%
where $f_\vw(\vx_i) = f_i$ is the function output at input $\vx_i$. The relationships specified are valid for generic likelihoods, and involve no approximations since the expectation is under the exact posterior, but given that the model can be expressed in a kernel formulation. \cref{eq:gp_pred} and \cref{eq:dual_param} highlight that the approximate inference technique, usually viewed as a posterior approximation, can be alternatively interpreted as an approximation of the expectation of gradients of loss/likelihoods.
